Standartox: A tool to derive exposure endpoints from chemical test results

### 2019-10-14 article/00_article.tex
Updates from Overleaf

\newcommand{\standartox}{STANDARTOX }
\newcommand{\epa}{EPA ECOTOX data base}
\newcommand{\app}{http://139.14.20.252:3838/etox-base-shiny/}
\newcommand{\git}{https://github.com/andreasLD/etox-base}
\newcommand{\gitapp}{https://github.com/andreasLD/etox-base-shiny}
\newcommand{\ecfifty}{EC\textsubscript{50}}

@@

\title{Etox-Base: A tool to derive exposure endpoints from chemical test results}

@@

1. Institute of Environmental Sciences, University of Koblenz-Landau Fortstraße 7, 76829 Landau, Germany {*}corresponding author(s):
Andreas Scharm{\"u}ller (scharmueller@uni-landau.de)

@@

\bibliographystyle{apalike}
\bibliography{refs/references-etox-base.bib,refs/references-etox-base-rpackages.bib,refs/references-etox-base-rshinypackages.bib}
% multiple refs have to be written w/o space

>>>>

\newcommand{\standartox}{STANDARTOX}
\newcommand{\epa}{EPA ECOTOX data base}
\newcommand{\rpackage}{http://139.14.20.252:3838/etox-base-shiny/}
\newcommand{\app}{www.standartox.uni-landau.de}
\newcommand{\git}{https://github.com/andreasLD/etox-base}
\newcommand{\gitapp}{https://github.com/andreasLD/etox-base-shiny}
\newcommand{\ecfifty}{EC\textsubscript{50}}

@@

\title{Standartox: A tool to derive exposure endpoints from chemical test results}

@@

1. Institute of Environmental Sciences, University of Koblenz-Landau Fortstraße 7, 76829 Landau, Germany {*}corresponding author(s):
Andreas Scharm{\"u}ller (andschar@protonmail.com)

@@

\bibliographystyle{apalike}
\bibliography{refs/references-standartox-build.bib,refs/references-standartox-app.bib}
% multiple refs have to be written w/o space

### 2019-10-14 article/parts/02_methods.tex
Updates from Overleaf

The \etoxbase{} consists of two parts of software: Firstly a processing pipeline (Fig. \ref{fig:pipeline}) downloads quarterly released new versions of the \epa{} and performs several preparation steps in R \citep{r_core_team_r_2017}, to then store the outcome in a PostgreSQL data base. This constitutes the basis upon which the R shiny web application is accessing the data \citep{r_core_team_r_2017}. The application itself is set of R functions that filter and aggregate the data according to a user's inputs \ref{fig:app}. Finally in the web application, a `filtered` and an `aggregated` data set as well as an interactive plot, for data exploration are returned. These two data sets can be downloaded. In the following these two parts are explained in detail, naming the respective R scripts in brackets.

@@

(bd\_epa\_download.R) 
(bd\_epa\_postgres.R)
(da\_epa1.R, da\_epa2.R, da\_epa3.R)
(qu\_aw.R)
(qu\_chemspieder\_scrape.R) 
(qu\_pc.R) 
(qu\_pp.R)
(qu\_worms2.R) 
(qu\_rgbif.R) 
(re\_merge.R) 
(re\_combine.R)
(re\_checks\_internal.R) 
(re\_analyses.R)
(re\_final.R)

The EPA ECOTOX data is downloaded and subsequently built into a PostgreSQL data base. Afterwards the data is imported into R for pre-processing, which includes validity checks, removal of special characters, unit harmonizations and selection of relevant variables. Chemical abstract service (CAS) numbers, unique chemical identifiers together we taxon names are then used to query other data bases for additional information on chemicals and biota respectively. Chemical information is retrieved from the Compendium of Pesticide Common Names \citep{CITE_AW}, the Chemspider data base \citep{CITE-CHEMSPIDER}, from Eurostat <CITATION>, the Pubchem data base \citep{CITE_PUBCHEM}, and from the Physprop data base \citep{CITE_PHYSPROP}. These queries mostly rely on the webchem R-package \citep{szocs_webchem_2015-1}. For biota, habitat and occurrence information is queried from the World Register of Marine Species (WORMS) \citep{WORMS} and from the Global Biodiversity Information Facility (GBIF) \citep{CITE_RGBIF} using the rgbif R-package \citep{chamberlain_rgbif_2018}. The acquired information is then merged and information is combined. From there on the data set is analyzed and checked and finally compiled in one table. This table is then used as the input for the filter and aggregation functions \ref{fig:pipeline}. For a detailed overview of what information is collected see table \ref{tab:scripts-pipeline}.

@@

@@

\subsection*{Aggregation}
The \etoxbase{} application aggregates the test results according to chosen filters in a two step process. Firstly the filtered test results are aggregated by the CAS number, the chosen taxon and the selected test duration. Secondly, the returned data is then aggregated by the CAS number. The former can't be influenced by the user and calculates either the minimum or the median depending on the amount of results to aggregate (n <= 2: minimum, if n > 2: median). Thereof the second step calculates the minimum, the maximum, the median, the geometric mean, or the arithmetic mean as an aggregate.

@@

The processing code is stored in a Github reopsitory (\git{}). All processing scripts were written in R version 3.4.4. PostgreSQL 9.5 has been used to build the data base. Hence, R together with its packages \ref{tab:rpackages} and PostgreSQL suffice to rebuild the process. The code of the application itself is stored in a second Github repostitory (\gitapp{}). Since new versions of the EPA ECOTOX data base are published on a quarterly basis the data acquisition process is automated and new versions of the application are scheduled to be released regularly.

>>>>

Standartox consists of three software parts which are explained in detail here. Firstly a processing pipeline (Fig. \ref{fig:pipeline}) downloads quarterly released new versions of the EPA ECOTOX data base and performs several preparation steps in R \citep{r_core_team_r_2017}, to then store the outcome in a PostgreSQL <CITE> data base. This constitutes the basis upon which the web application and and application programming interface (API) together with a R package are accessing the data \citep{r_core_team_r_2017}.

\subsection{}

The application itself is set of R functions that filter and aggregate the data according to a user's inputs \ref{fig:app}. Finally in the web application, a `filtered` and an `aggregated` data set as well as an interactive plot, for data exploration are returned. These two data sets can be downloaded. In the following these two parts are explained in detail, naming the respective R scripts in brackets.

@@

The EPA ECOTOX data is downloaded and subsequently built into a PostgreSQL data base. Afterwards the data is imported into R for pre-processing, which includes validity checks, removal of special characters, unit harmonizations and selection of relevant variables. Chemical abstract service (CAS) numbers, unique chemical identifiers together with taxon names are then used to query other data bases for additional information on chemicals and organisms respectively. Chemical information is retrieved inter alia from the Compendium of Pesticide Common Names \citep{CITE_AW}, the Chemspider data base \citep{CITE-CHEMSPIDER}, from Eurostat <CITATION>, the Pubchem data base \citep{CITE_PUBCHEM}, and from the Physprop data base \citep{CITE_PHYSPROP}. These queries mostly rely on the webchem R-package \citep{szocs_webchem_2015-1}. For biota, habitat and occurrence information is queried from the World Register of Marine Species (WORMS) \citep{WORMS} and from the Global Biodiversity Information Facility (GBIF) \citep{CITE_RGBIF} using the rgbif R-package \citep{chamberlain_rgbif_2018}. The acquired information is then merged and information is combined. From there on the data set is analyzed and checked and finally compiled in one table. This table is then used as the input for the filter and aggregation functions \ref{fig:pipeline}. For a detailed overview of what information is collected see table \ref{tab:scripts-pipeline}.

@@

\subsubsection*{Aggregation}
The \etoxbase{} application aggregates the test results according to chosen filters in a two step process. Firstly the filtered test results are aggregated by the CAS number, the chosen taxon and the selected test duration. Secondly, the returned data is then aggregated by the CAS number. The former can't be influenced by the user and calculates either the minimum or the median depending on the amount of results to aggregate (n <= 2: minimum, if n > 2: median). Thereof the second step calculates the minimum, the maximum, the median, the geometric mean, or the arithmetic mean as an aggregate.

\subsubsection*{Future of the project}
Since new versions of the EPA ECOTOX data base are published on a quarterly basis the data acquisition process is automated and new versions of the Standartox are scheduled to be released regularly.

The code for the web-application and the application programming interface are placed here: https://github.com/andschar/standartox-app

@@

\subsection*{The R package}

The code for the R-package standartox is placed here: https://github.com/andschar/standartox

@@

The processing pipeline code is stored here https://github.com/andschar/standartox-build, the code for the web application is stored under https://github.com/andschar/standartox-app and the R-pckage code is stored in https://github.com/andschar/standartox. The code can freely be accessed under the MIT License <CITATION>. Most of the code is written in R 3.6.1 and some parts in PostgreSQL 9.6.1. An exhaustive list of additional R-pakage used can be found in the supplementary material <REFERENCE-TO-SUPPL>.
